Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-10-11 main.tex
include proofs in text. rename files.

%\subsection{Properties}
\medskip\noindent
Similarly to \VPA~\cite{AlurMadhusudan09nested}
and \SVPA~\cite{dAntonyAlur14SVPDA},
the class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}\label{prop:SWVPA-product}
Let $A_1$ and $A_2$ be two \SWVPA
over the same $\Delta$, commutative $\Semiring$ and effective $\bar\Phi$.
There exist two effectively constructible $\SWVPA$
$A_1 \oplus A_2$ and $A_1 \otimes A_2$,
such that for every $s \in \Delta^*$,
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
%
\begin{proof}
We do a classical product construction, see Appendix~\ref{sec:product} for details.
%similar to the case of the Boolean semiring~\cite{dAntonyAlur14SVPDA}.
\end{proof}
\noindent
%We present now a procedure for searching, for a~$\SWVPA$ $A$,
%a word %$w \in \Delta^*$
%of minimal weight for~$A$.
We present now a procedure for searching a word of minimal weight for a~$\SWVPA$~$A$.
%as stated in the following proposition. %\wrt~$\leq_\oplus$,
%It requires the following property of label theories.
%
\begin{proposition}\label{th:best-search}
For a \SWVPA $A$
over $\Delta$,
$\Semiring$ commutative, bounded, total and complete, %semiring
and $\bar\Phi$ effective, %and $k$-convex, % label theory,
one can construct in PTIME a word $t \in \Delta^*$
such that $A(t)$ is minimal \wrt the natural ordering $\leq_\oplus$ for $\Semiring$.
\end{proposition}
%\florent{total?}
%
Let $A = \< Q, P, \init, \bar\wei, \final >$.
%
We propose a Dijkstra-like algorithm computing,
for every $q, q' \in Q$,
the minimum, \wrt $\leq_\oplus$, of the function
$\beta_{q, q'} : t \mapsto \weight_A(\config{q}{\bot}, t, \config{q'}{\bot})$.
Let us denote by $b_\bot(q, q')$ this minimum.
By definition of $\leq_\oplus$, and since $\Semiring$ is total,
\reviews{1) $\Semiring$ total $\to$ $\Semiring$ extremal}
it holds that:
%
\begin{equation}\label{eq:bbot}
  b_\bot(q, q') = \bigoplus_{t\in \Delta^*}
  \textstyle
  \weight_A\bigl(\config{q}{\bot}, t, \config{q'}{\bot}\bigr).
\end{equation}
The infinite sum in~\eqref{eq:bbot} is well defined since $\Semiring$ is complete.
%
Following~\eqref{eq:SWVPA-value}, and the associativity, commutativity
and distributivity for~$\otimes$ and~$\oplus$, the minimum of~$A(t)$ is:
\begin{equation}\label{eq:min}
{\displaystyle\bigoplus_{t\in \Delta^*}} A(t)
=
{\displaystyle\bigoplus_{t\in \Delta^*}}
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
\beta_{q, q'}(t)
\mathop{\otimes} \mathsf{out}(q')
=
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
b_\bot(q, q')
\mathop{\otimes} \mathsf{out}(q')
\end{equation}

@@

\input{altalgo}

\begin{algorithm}
\SetKw{Extract}{extract}
\SetKw{Update}{update}

\textbf{initially} let $\Q = (Q \times Q) \cup (Q \times P \times Q)$,
and let $d_\bot(q_1, q_2) = d_\top(q_1, p, q_2) = \one$
if $q_1 = q_2$ and $d_\bot(q_1, q_2) = d_\top(q_1, p, q_2) = \zero$ otherwise$\;$

\smallskip\noindent
\While{$\Q \neq \emptyset$}{

%\noindent\quad
\Extract $\< q_1, q_2>$ or $\< q_1, p, q_2>$ from $\Q$
%
%\noindent\quad
such that $d_\bot(q_1, q_2)$, 
resp. $\bigoplus_{\Deltac} d_\top(q_1, p, q_2)$, is minimal in $\Semiring$ \wrt $\leq_\oplus$$\;$

%\noindent\quad
%\textbf{for all} $q_0, q_3 \in Q$
\Update $d_\bot$ if $\< q_1, q_2>$ was extracted (Figure~\ref{fig:best-update-bottom}), or \\
\Update $d_\top$ if $\< q_1, p, q_2>$ was extracted (Figure~\ref{fig:best-update-top}).$\;$
}
\caption{Best search for \SWVPA}
\label{algo:Dijkstra}
\end{algorithm}

\input{proof}

Algorithm~\ref{algo:Dijkstra} constructs iteratively, using a priority queue $\Q$, 
two markings
$d_\bot : Q \times Q \to \Semiring$ and
$d_\top : Q \times P \times Q \to \Phic$,
%of the triplets $\<q, \sigma, q'>$
%of states of $A$ by weight values in $\Semiring$,
that converge eventually to $b_\top$ and $b_\bot$. 
%
\florent{intuition sur update en commentaire}
%Intuitively, in the update defined in Figure~\ref{fig:best-update},
%we assume that the values of $d_\bot$ and $d_\top$ are definitely 
%fixed for every tuplet which has been extracted from the queue $\Q$ 
%(like in particular $\<q_1, q_2>$ or $\< q_1, p, q_2>$ when the algorithm calls this update), 
%and we use these values 
%(and the functions $w$ of transitions) 
%in order to update the values of $d_\bot$ and $d_\top$ 
%on the other tuplets (those which are still in the queue $\Q$).
%
\florent{explication Fig.~\ref{fig:best-update} (update) suivant cas de \eqref{eq:SWVPA-weight}?}
\florent{The infinite sums in the updates of $d$ in Algorithm~\ref{algo:Dijkstra},
 Figure~\ref{fig:best-update} are well defined since~$\Semiring$ is complete.}
%
% COMPLEXITY
%The algorithm performs $2.|Q|^2$ iterations until $P$ is empty,
%and each iteration has a time complexity $O(|Q|^2 . |P|)$.
%\florent{complexity: detail with and states?}
%That gives a time complexity $O(|Q|^4 . |P|)$.
%It can be reduced by implementing $P$ as a priority queue,
%prioritized by the value returned by $d$.
%** where  $T$ is the worst time complexity for an oracle computing the values
%of Definition~\ref{def:effective}.
%$|Q|^3.\log(|Q|^2)$
%
%Algorithm~\ref{algo:Dijkstra}
It terminates in PTIME (when $\Q$ is empty), 
and its correctness is ensured by~\eqref{eq:min} and  
the following invariants: 
$\< q_1, q_2> \notin \Q$ iff $d_\bot(q_1, q_2) = b_\bot(q_1, q_2)$, 
and $\< q_1, p, q_2> \notin \Q$ iff 
$\bigoplus_{\Deltac} d_\top(q_1, p, q_2) = \bigoplus_{\Deltac} b_\top(q_1, p, q_2)$.
\florent{(see Appendix~\ref{sec:bestsearch}).}
%of Algorithm~\ref{algo:Dijkstra}.

\begin{figure}[ht]
\(
\begin{array}{lcl}
\multicolumn{3}{l}{%
 \mbox{For~every}\  q_0, q_3 \in Q,}\\[1mm] % if $\< q_1, p, q_2> \notin \Q$
%
d_\bot(q_0, q_3) & \opluseq &
  d_\bot(q_0, q_2) \otimes
  \bigoplus_{\Deltai} \weiei(q_2, q_3)\ 
  \mathrm{if~} \< q_0, q_2> \notin \Q \\
%
d_\bot(q_0, q_3) & \opluseq &
  {\bigoplus_{\Deltac}}
  \bigl(
  \weiec(q_0, p, q_1) \otimes
   d_\top(q_1, p, q_2) \otimes
  \bigoplus^2_{\Deltar} \weir(q_2, p, q_3)\bigr)\ 
  \mathrm{if~} \< q_1, p, q_2> \notin \Q \\
%
d_\bot(q_0, q_3) & \opluseq &
  d_\bot(q_0, q_2) \otimes
  \bigoplus_{\Deltar} \weier(q_2, q_3)\ 
  \mathrm{if~} \< q_0, q_2> \notin \Q \\ 
%
d_\bot(q_0, q_3) & \opluseq &
  d_\bot(q_0, q_2) \otimes d_\bot(q_2, q_3)\  
  \mathrm{if~} \< q_0, q_2>\notin \Q, \< q_2, q_3> \notin \Q \\[-1mm]
\end{array}
\)
\caption{\textbf{update} $d_\bot$ (Algorithm~\ref{algo:Dijkstra}).}
\label{fig:best-update-bottom}
\end{figure}
%

\begin{figure}[ht]
\(
\begin{array}{lcl}
\multicolumn{3}{l}{%
 \mbox{For~every}\  q_0, q_3 \in Q\ \mbox{and}\ p \in P,}\\[1mm] % if $\< q_1, p, q_2> \notin \Q$
d_\top(q_0, p, q_3) & \opluseq &
  d_\top(q_0, p, q_2) \otimes
  \bigoplus_{\Deltai} \weii(q_2, p, q_3)\
  \mathrm{if~} \< q_0, p, q_2> \notin \Q \\
%\multicolumn{3}{l}{\quad
%\mathrm{where~} \weii(q_2, p, q_3)_a \in \Phic
%\mathrm{~is~the~partial~application~}
%x_c \mapsto \weii(q_2, x_c, p, a, q_3)}\\
d_\top(q_0, p, q_3) & \opluseq &
  {\bigoplus^2_{\Deltac}}
  \bigl[ \bigl( \weic(q_0, p, p', q_1) \otimes_2
  d_\top(q_1, p', q_2) \bigr) \otimes_2
  {\bigoplus_{\Deltar}} \weir(q_2, p', q_3) \bigr]\
  \mathrm{if~} \< q_1, p', q_2> \notin \Q \\
%\multicolumn{3}{l}{\quad
%\mathrm{where~} \weic(q_0, p, p', q_1)_{c'} \in \Phic
%\mathrm{~is~the~partial~application~}
%x_c \mapsto \weic(q_0, p, x_c, c', p', q_1)}\\[2pt]
%
d_\top(q_0, p, q_3) & \opluseq &
  d_\top(q_0, p, q_2) \otimes d_\top(q_2, p, q_3)\ 
  \mathrm{if~} \< q_0, p, q_2> \notin \Q, \< q_2, p, q_3> \notin \Q \\[-1mm]
\end{array}
\)
%$\weii(q_2, p, q_3)_a \in \Phic$, for $a \in \Deltai$,
%is the partial application
%$x_c \mapsto \weii(q_2, x_c, p, a, q_3)$.
%
%$\weic(q_0, p, p', q_1)_{c'} \in \Phic$, for $c' \in \Deltac$,
%is the partial application
%$x_c \mapsto \weic(q_0, p, x_c, c', p', q_1)$.
%\caption{Update $d_\bot$ with $\<q_1, q_2>$ or $d_\top$ with $\< q_1, p, q_2>$.}
\caption{\textbf{update} $d_\top$ (Algorithm~\ref{algo:Dijkstra}).}
\label{fig:best-update-top}
\end{figure}
%
%\noindent
%** effectively computable by hypothesis that the label theory is effective**
%\florent{REVOIR}
Thanks to the hypothesis that $\bar\Phi$ is effective, 
it is possible to construct during the iteration a witness word
for Proposition~\ref{th:best-search}, 
\ie a word $t \in \Delta^*$ with a minimal weight $A(t)$ \wrt~$\leq_\oplus$.

%we require the additional property of convexity of weight functions.***
%using the second requirement in Definition~\ref{def:effective}.

@@

\section{Properties of Label Theory Operators} \label{sec:closure}
\input{app-labels}

@@

\section{Proof of Proposition~\ref{prop:SWVPA-product}} \label{sec:product}
\input{app-product}

@@

\section{Nested Words and Parse Trees} \label{sec:trees}\label{app:trees}
\input{app-trees}

>>>>

%\subsection{Properties}
\medskip\noindent
Similarly to \VPA~\cite{AlurMadhusudan09nested}
and \SVPA~\cite{dAntonyAlur14SVPDA},
the class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}\label{prop:SWVPA-product}
Let $A_1$ and $A_2$ be two \SWVPA
over the same $\Delta$, commutative $\Semiring$ and effective $\bar\Phi$.
There exist two effectively constructible $\SWVPA$
$A_1 \oplus A_2$ and $A_1 \otimes A_2$,
such that for every $s \in \Delta^*$,
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
%
\begin{proof}
We do a classical product construction. % see Appendix~\ref{sec:product} for details.
%similar to the case of the Boolean semiring~\cite{dAntonyAlur14SVPDA}.
\input{app-product}
\end{proof}
\noindent
%We present now a procedure for searching, for a~$\SWVPA$ $A$,
%a word %$w \in \Delta^*$
%of minimal weight for~$A$.
We present now a procedure for searching a word of minimal weight for a~$\SWVPA$~$A$.
%as stated in the following proposition. %\wrt~$\leq_\oplus$,
%It requires the following property of label theories.
%
\begin{proposition}\label{th:best-search}
For a \SWVPA $A$
over $\Delta$,
$\Semiring$ commutative, bounded, total and complete, %semiring
and $\bar\Phi$ effective, %and $k$-convex, % label theory,
one can construct in PTIME a word $t \in \Delta^*$
such that $A(t)$ is minimal \wrt the natural ordering $\leq_\oplus$ for $\Semiring$.
\end{proposition}
%\florent{total?}
%
Let $A = \< Q, P, \init, \bar\wei, \final >$.
%
We propose a Dijkstra-like algorithm computing,
for every $q, q' \in Q$,
the minimum, \wrt $\leq_\oplus$, of the function
$\beta_{q, q'} : t \mapsto \weight_A(\config{q}{\bot}, t, \config{q'}{\bot})$.
Let us denote by $b_\bot(q, q')$ this minimum.
By definition of $\leq_\oplus$, and since $\Semiring$ is total,
\reviews{1) $\Semiring$ total $\to$ $\Semiring$ extremal}
it holds that:
%
\begin{equation}\label{eq:bbot}
  b_\bot(q, q') = \bigoplus_{t\in \Delta^*}
  \textstyle
  \weight_A\bigl(\config{q}{\bot}, t, \config{q'}{\bot}\bigr).
\end{equation}
The infinite sum in~\eqref{eq:bbot} is well defined since $\Semiring$ is complete.
%
Following~\eqref{eq:SWVPA-value}, and the associativity, commutativity
and distributivity for~$\otimes$ and~$\oplus$, the minimum of~$A(t)$ is:
\begin{equation}\label{eq:min}
{\displaystyle\bigoplus_{t\in \Delta^*}} A(t)
=
{\displaystyle\bigoplus_{t\in \Delta^*}}
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
\beta_{q, q'}(t)
\mathop{\otimes} \mathsf{out}(q')
=
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
b_\bot(q, q')
\mathop{\otimes} \mathsf{out}(q')
\end{equation}

@@

%\input{algo-best}
\input{algo-best-alt}

@@

%\section{Properties of Label Theory Operators} \label{sec:closure}
%\input{app-labels}

@@

%\section{Proof of Proposition~\ref{prop:SWVPA-product}} \label{sec:product}
%\input{app-product}

@@

%\section{Nested Words and Parse Trees} \label{sec:trees}\label{app:trees}
%\input{app-trees}
